---
title: "Which storm events are most harmful to population health and have the greatest economic consequences"
output: html_document
---
##Sinopsys

This study aims to identify the types of events that are most harmful to the population health or have the greatest economic consequences. This report is based on data from events of a broad variety registered between 1950 and 2011. The analysis explores the total number of reported victims related to events as well as the amount of losses with damage on economic goods. 

## Data processing
### Loading Raw Data
The data used was obtained from the National Oceanic and Atmospheric Administration's (NOAA) [National Weather Service](http://www.nws.noaa.gov/directives/). The [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)[47 MB] documents "...the occurrence of storms and other significant weather phenomena having
sufficient intensity to cause loss of life, injuries, significant property damage, and/or disruption to commerce", according to National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

#### Environment setting

```{r echo=TRUE, results="hide", message=FALSE, warning=FALSE}
## load needed libraries
library(dplyr)
library(lattice)
library(ggplot2)
library(grid)
library(R.utils)
options(scipen = 1)
```

#### Reading in data

```{r echo=TRUE, cache=TRUE}
## setting vaviable 
setInternet2(use = TRUE)
urlData <-
        "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName <- "repdata-data-StormData.csv.bz2"
destName <- "repdata-data-StormData.csv"
## Download the file if it isn't yet downloaded
if (!file.exists(fileName)) {
        download.file(url = urlData, destfile = fileName, mode = "wb")
}
if (!file.exists(destName)) {
        bunzip2(fileName, destName, overwrite = TRUE, remove = FALSE)
}
## read the file if it isn't yet read
if (!"stormData" %in% ls()) {
        stormData <- read.csv(destName)
}
## check the amount of data
dim(stormData)
head(stormData, 3)
```

We can see that there is a few variables of interest to the scope of this analysis. We may now subset the dataset to get just the list of interest.  

```{r echo=TRUE, cache=TRUE}
## We need now to adjust the column names
names(stormData) <- tolower(names(stormData))
healthdata <- select(stormData, evtype, fatalities, injuries)
damagedata <- select(stormData, evtype, propdmg, propdmgexp, cropdmg, cropdmgexp)
```

## Results

### Across the United States, which types of events are most harmful with respect to population health? 

To evaluate the danger of each event to the population health, we calculate the deaths and injuries caused by each event type, then we  filter the 10 worst.

```{r echo=TRUE}
totalcases <- aggregate(.~ evtype, data = healthdata, FUN = sum)
summary(totalcases)
fatalmost <- totalcases[order(-totalcases$fatalities),][1:10,]
injurmost <- totalcases[order(-totalcases$injuries),][1:10,]
fatalmost[,c("evtype", "fatalities")]
injurmost[,c("evtype", "injuries")]
```

The following graph shows the comparison on the number of fatalities and injuries caused by the 10 worst weather events to the population health.

```{r echo=TRUE}
par(mar = c(6,4,4, 2), mfrow = c(1, 2))
barplot(
        fatalmost$fatalities,
        names.arg = fatalmost$evtype,
        las = 2,
        cex.names = 0.5,
        ylim = c(0, 6000),
        main = "Fatalities caused by \n Severe Weather Events \n in U.S. (1950 - 2011)",
        ylab = "Number of Fatalities"
)
barplot(
        injurmost$injuries / 1000,
        names.arg = injurmost$evtype,
        las = 2,
        cex.names = 0.5,
        ylim = c(0, 100),
        main = "Injuries caused by \n Severe Weather Events \n in U.S. (1950 - 2011)",
        ylab = "Number of Injuries(in thousand cases)"
)

```

### Across the United States, which types of events have the greatest economic consequences?

To evaluate the damages caused by each event to the economy, we calculate the property damages and crop damages, caused by each event type, then we  filter the 10 worst.

```{r echo=TRUE, warning=FALSE}
## we need to adjust some values to avoid errors
adjustvalues <- function(dmg,exp){
        if (is.na(dmg) || is.null(dmg)) {return(0)}
        if (is.na(exp) || is.null(exp)) {return(ifelse(is.numeric(dmg),dmg,0))}
        if (toupper(exp)=='B') {return(dmg*10^9)}
        if (toupper(exp)=='M') {return(dmg*10^6)}
        if (toupper(exp)=='K') {return(dmg*10^3)}
        if (toupper(exp)=='H') {return(dmg*10^2)}
        if (exp=='0') {return(dmg)}
        return(dmg)
}

damagedata$losses <- adjustvalues(damagedata$propdmg, damagedata$propdmgexp) + adjustvalues(damagedata$cropdmg, damagedata$cropdmgexp)

## compute the sum of losses by event type
totalloss <- aggregate(losses ~ evtype, data = damagedata, FUN = sum)
summary(filter(totalloss, losses > 0))
damagemost <- totalloss[order(-totalloss$losses),][1:10,]
damagemost
```

The following graph shows the comparison on the economic consequences caused by the 10 worst weather events.


```{r echo=TRUE}
par(mar = c(6,4,4, 2))
barplot(
        damagemost$losses / 1000000,
        names.arg = damagemost$evtype,
        las = 2,
        cex.names = 0.5,
        main = "Damages caused by \n Severe Weather Events \n in U.S. (1950 - 2011)",
        ylim = c(0, 3500),
        ylab = "Damages (in Million US$)"
)
``` 

## Conclusion

Tornados are the most harmful weather events to popullation healt as well as to the economy.
